---
title: "Correlation (Field Ch6)"
author: "Kyla McConnell"
date: "10/21/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(corrplot) #for correlation matrices
library(PerformanceAnalytics) #for correlation matrices
library(GGally) #for correlation matrices
library(psych) #for correlation matrices

#Exam data: shows the exam scores (Exam), anxiety levels (Anxiety), and amount of time preparing (Revise)
examData <- read.delim("~/Documents/PhD/Coding/R/Field_DiscStatswithR/Data/ExamAnxiety.txt", header=TRUE) 
#update this with the file location in your directory (you can find it on CH6 of Field's website, I had trouble so had to use an absolute path)
examData2 <- examData[, c("Exam", "Anxiety", "Revise")] #includes only the interval variables
```


# Correlation matrices

## corrplot
Package: corrplot
Documentation: https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html

Plots the correlation between many variables at the same time in a matrix.

Form: corrplot(cor output, type="string", method="string", tl.cex=0.9)
cor output: first conduct cor() and save it to an item
type: "full" for full matrix (where every combination appears twice), or "upper" or "lower" for one half
method: "number" for (colored) correlaton coefficients, can also use "circle", "square", "ellipse", "shade", "color" or "pie" (but these are often confusing)
upper: set upper="pie/etc" to set the upper half to a certain look
lower: set lower="number/etc" to set the lower half to a certain look (so you can combine two different display modes)
order: see documentation, recommendation is probably "hclust"

See documentation for more, incl. coloring, labels, etc.

```{r}
exam_cor = cor(examData2) #creates correlation item
corrplot(exam_cor, type="upper",  method = "number", tl.cex = 0.9)#order="hclust",

```

## chart.Correlation
Package: PerformanceAnalytics
Documentation: https://www.rdocumentation.org/packages/PerformanceAnalytics/versions/1.5.3/topics/chart.Correlation

Plots the correlation between many variables at the same time in a matrix, plus the result of cor.test() as stars, and an optional histogram.

Form: corrplot(as.matrix(data), histogram=TRUE, method="string")
data: must be a matrix or converted to a matrix
histogram: TRUE/FALSE
method: "spearman", "pearson", "kendall", or multiple c("spearman", "pearson")

NOTE: Significance stars are for your information but should not be relied upon because the p-values should be Bonferonni-corrected first.

```{r}
chart.Correlation(as.matrix(examData2),  method = "spearman")

```

## ggpairs
Package: GGally
Documentation: https://www.rdocumentation.org/packages/GGally/versions/1.4.0/topics/ggpairs

Form: ggpairs(data, upper= list(continuous=wrap('cor', method="spearman")))

data: Dataset of variables
upper: Plots the correlation coefficients of the given method using the cor plot, continuous meaning that the variables are continuous.

See documentation for the rest of the very complicated form.

```{r}
ggpairs(examData2, upper = list(continuous = wrap('cor', method = "spearman")))

```

## pairs.panels
Package: psych
Documentation: https://www.rdocumentation.org/packages/psych/versions/1.8.12/topics/pairs.panels

Form: pairs.panels(data, scale=TRUE, method="string")
data: dataset
scale: TRUE or FALSE
method: "spearman", "pearson", "kendall"

See documentation for more.

```{r}
pairs.panels(examData2, scale=TRUE, method = "spearman")

```

